---
title: "DS analysis"
author: 
- name: Helena L. Crowell
  affiliation: 
  - &IMLS Institute for Molecular Life Sciences, University of Zurich, Switzerland
  - &SIB Swiss Institute of Bioinformatics (SIB), University of Zurich, Switzerland
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE)
options(width = 100)
```

### load packages
```{r load-libs, warning = FALSE, message = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(dplyr)
library(limma)
library(magrittr)
library(muscat)
library(scater)
library(SingleCellExperiment)
```

# Data preparation

## Load data & prep. for `muscat`

```{r prep-data}
# load data
sce <- readRDS(file.path("output", "MAGL-SCE.rds"))

# make WT reference group
sce$group_id <- factor(sce$group_id, levels = c("WT", "LPS"))

# reorder sample levels
m <- match(levels(sce$sample_id), sce$sample_id)
o <- order(sce$group_id[m])
sce$sample_id <- factor(sce$sample_id, levels = levels(sce$sample_id)[o])

# prep. SCE for `muscat`
sce <- prepSCE(sce, "cluster_id", "sample_id", "group_id")
(ei <- metadata(sce)$experiment_info)
```

## Cluster annotation

In consideration of the visualizations provided in the [Annotation](docs/2-annotation.html) tab and additional exploration with `r Biocpkg("iSEE")`, we arrive at the following cluster annotations: 

```{r cluster-anno, warning = FALSE}
anno <- list(
    "Astrocytes" = 3,
    "Endothelial" = 13,
    "Microglia" = 17,
    "Oligodendrocytes" = 4,
    "OPC" = 12,
    "Excit_Neuron" = c(0, 1, 2, 5, 6, 7, 10, 14, 16),
    "Inhib_Neuron" = c(8, 9, 11, 15),
    "CPE cells" = c(18, 19))

m <- match(sce$cluster_id, unlist(anno))
ns <- vapply(anno, length, numeric(1))
lab <- rep.int(names(anno), ns)
sce$cluster_id <- lab[m]

# cluster-sample cell-counts
table(sce$cluster_id, sce$sample_id)

# normalize for visualization
sce <- normalize(sce)
```

# Aggregation of single-cell to pseudo-bulk data

```{r aggregation}
system.time(pb <- aggregateData(sce))
pb
```

## Pseudobulk-level MDS plot

Prior to conducting any formal testing, we can compute a multi-dimensional scaling (MDS) plot of aggregated signal to explore overall sample similarities. Ideally, such a represenation of the data should separate both clusters and groups from one another. Vice versa, samples from the same cluster/group should fall close to each other.

In our MDS plot on pseudobulk counts (Fig. \@ref(fig:pb-mds)), we can observe that cell-populations (clusters) are separated quite well, while WT and stimulated samples (groups) are separated to a lesser (if any) extend.

```{r pb-mds, fig.wide = TRUE, fig.cap = "Pseudobulk-level MDS plot. Each point represent one cluster-sample instance;\npoints are colored by cluster ID and shaped by group ID."}
pbMDS(pb)
```

# DS analysis

```{r ds-analysis}
system.time(res <- pbDS(pb, verbose = FALSE))
tbl <- bind_rows(res$table[[1]])
tbl <- filter(tbl, p_adj.loc < 0.05, abs(logFC) > 1)
tbl <- mutate_if(tbl, is.numeric, format, digits = 4)
```

# Save SCE & results

```{r save-rds}
saveRDS(sce, file.path("output", "MAGL-SCE.rds"))
saveRDS(res, file.path("output", "MAGL-DS_res.rds"))
write.csv(tbl, file.path("output", "MAGL-DS_res.csv"), row.names = FALSE)
```