---
title: "DS analysis"
author: 
- name: Helena L. Crowell
  affiliation: 
  - &IMLS Institute for Molecular Life Sciences, University of Zurich, Switzerland
  - &SIB Swiss Institute of Bioinformatics (SIB), University of Zurich, Switzerland
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
    code_folding: hide
---

```{r include = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE)
options(width = 100)
```

### load packages
```{r warning = FALSE, message = FALSE}
library(ComplexHeatmap)
library(cowplot)
library(limma)
library(magrittr)
library(muscat)
library(scater)
library(SingleCellExperiment)
```

# Data preparation

## Load data & prep. for `muscat`

```{r prep-data}
# load data
sce0 <- readRDS(file.path("output", "MAGL-SCE_tmp.rds"))

# concat. sample & group ID
sce0$id <- factor(paste(sce0$sample_id, sce0$group_id, sep = "_"))
# make WT reference group
sce0$group_id <- factor(sce0$group_id, levels = c("WT", "LPS"))
# reorder sample levels
o <- sapply(levels(sce0$group_id), grep, levels(sce0$id))
sce0$id <- factor(sce0$id, levels = levels(sce0$id)[o])

# prep. SCE for `muscat`
sce <- prepSCE(sce0,
    group_id = "group_id",
    sample_id = "id",
    cluster_id = "integrated_snn_res.0.2",
    drop = FALSE)

(ei <- metadata(sce)$experiment_info)
```

## Cluster annotation

In consideration of the visualizations provided in the [Annotation](docs/2-annotation.html) tab and additional exploration with `r Biocpkg("iSEE")`, we arrive at the following cluster annotations: 

```{r cluster-anno, warning = FALSE}
anno <- list(
    "excit_neuron1" = c(0, 1, 4, 5, 16),
    "excit_neuron2" = 7,
    "excit_neuron3" = c(2, 20),
    "excit_neuron4" = 9,
    "excit_neuron5" = 14,
    "excit_neuron6" = 8,
    "inhib_neuron1" = c(10, 15),
    "inhib_neuron2" = 11,
    "inhib_neuron3" = 12,
    "Astrocytes" = 3,
    "OPC" = 13,
    "Oligodendrocytes" = 6,
    "Endothelial" = 18,
    "Microglia" = 19)
# unassigned: 17, 21, 22, 23

m <- match(sce$cluster_id, unlist(anno))
ns <- vapply(anno, length, numeric(1))
lab <- rep.int(names(anno), ns)
sce$cluster_id <- lab[m]

# remove unassigned cells
sce <- sce[, !is.na(sce$cluster_id)]
sce$cluster_id <- factor(sce$cluster_id)

# cluster-sample cell-counts
table(sce$cluster_id, sce$sample_id)

# normalize for visualization
sce <- normalize(sce)
```

# Aggregation of single-cell to pseudo-bulk data

```{r agg}
system.time(pb <- aggregateData(sce))
# SCE of dim. #(genes) x #(samples)
pb 
# one assay per cluster
assayNames(pb) 
# view 1st cluster
head(assay(pb))
```

## Pseudobulk-level MDS plot

Prior to conducting any formal testing, we can compute a multi-dimensional scaling (MDS) plot of aggregated signal to explore overall sample similarities. Ideally, such a represenation of the data should separate both clusters and groups from one another. Vice versa, samples from the same cluster/group should fall close to each other.

In our MDS plot on pseudobulk counts (Fig. \@ref(fig:pb-mds)), we can observe that cell-populations (clusters) are separated quite well, while WT and stimulated samples (groups) are separated to a lesser (if any) extend.

```{r pb-mds, fig.wide = TRUE, fig.cap = "Pseudobulk-level MDS plot. Each point represent one cluster-sample instance;\npoints are colored by cluster ID and shaped by group ID."}
pbMDS(pb)
```

# DS analysis

```{r ds-analysis}
# specify design & contrast matrix
(design <- model.matrix(~ 0 + ei$group_id) %>% 
    set_rownames(ei$sample_id) %>% 
    set_colnames(levels(ei$group_id)))
(contrast <- makeContrasts("LPS-WT", levels = design))
system.time(res <- pbDS(sce, pb, design, contrast, verbose = FALSE))
```

# Results filtering & overview

To get a general overview of the differential testing results, we first filter them to retain hits with $\text{FDR}<5\%$ and $~|~logFC~|~>1$, and view the number & percentage of differential findings by cluster. Finally, we extract the top hits (lowest adj. p-value) in each cluster.

```{r}
tbl0 <- res$table$`LPS-WT`
# filter abs(logFC) > 1 & FDR < 0.05
tbl <- lapply(tbl0, dplyr::filter, abs(logFC) > 1, p_adj.loc < 0.05)
# sort by FDR
tbl <- lapply(tbl, dplyr::arrange, p_adj.loc)

# nb. & % of hits by cluster
n_de <- vapply(tbl, nrow, numeric(1))
cbind(n_de, perc = round(n_de / nrow(sce) * 100, 2))

# get top-hits for ea. cluster
top_gs <- lapply(tbl, function(u) {
    n <- ifelse(nrow(u) < 6, nrow(u), 6)
    u$gene[seq_len(n)]
})
```

# Visualization results

## Cell-level visualization

The `r Biocpkg("scater")`[@McCarthy2017-scater] packages provides a variety of visualizations for single-cell data. Here, we use `plotExpression` to render violin plots of the top differential genes identified for each cluster. We specify `x = "sample_id"` to obtain one violin per sample, and `colour_by = "group_id"` to signify the experimental condition each sample belongs to. 

```{r violins, collapse = TRUE}
# split cells by cluster
cs_by_k <- split(colnames(sce), sce$cluster_id)

kids <- levels(sce$cluster_id)
names(kids) <- kids

ps <- lapply(kids, function(k) 
  if (length(top_gs[[k]]) != 0)
    plotExpression(sce[, cs_by_k[[k]]], features = top_gs[[k]], 
        x = "sample_id", colour_by = "group_id") +
        guides(fill = guide_legend(override.aes = list(size = 2, alpha = 1))) +
        theme_classic() + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))) 
```

### &nbsp; {.tabset}
```{r echo = FALSE, results = "asis"}
for (k in kids) {
  cat("#### ", k, "\n")
  print(ps[[k]])
  cat("\n\n")
}
```

## Sample-level visualization {.tabset}

```{r diff-hm, warning = FALSE, collapse = TRUE}
# normalize & compute cluster-sample expression-means
ms <- aggregateData(sce, assay = "logcounts", fun = "mean")

# wrapper for z-normalization
.z_norm <- function(x, th = 2.5) {
    x <- as.matrix(x)
    sds <- rowSds(x, na.rm = TRUE)
    sds[sds == 0] <- 1
    x <- t(t(x - rowMeans(x, na.rm = TRUE)) / sds)
    #x <- (x - rowMeans(x, na.rm = TRUE)) / sds
    x[x >  th] <-  th
    x[x < -th] <- -th
    return(x)
}

# wrapper to plot differential heatmap
.plot_diff_hm <- function(ms, k, gs, ei) {
    mat <- .z_norm(assays(ms)[[k]][gs, , drop = FALSE])
    m <- match(colnames(mat), ei$sample_id)
    cols <- scales::hue_pal()(nlevels(ei$group_id))
    names(cols) <- levels(ei$group_id)
    col_anno <- columnAnnotation(
        df = data.frame(group_id = ei$group_id[m]),
        col = list(group_id = cols),
        gp = gpar(col = "white"),
        show_annotation_name = FALSE)
    Heatmap(mat, 
        column_title = k,
        name = "z-normalized\nexpression",
        col = c("royalblue", "cornflowerblue", "white", "gold", "orange"),
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_side = "left",
        row_names_gp = gpar(fontsize = 6),
        rect_gp = gpar(col = "white"),
        top_annotation = col_anno)
}
hms <- lapply(kids, function(k)
  if (nrow(tbl[[k]]) != 0) {
    n <- min(nrow(tbl[[k]]), 10)
    .plot_diff_hm(ms, k, tbl[[k]]$gene[seq_len(n)], ei)
  })
```

```{r echo = FALSE, results = "asis"}
for (k in kids) {
  cat("### ", k, "\n")
  print(hms[[k]])
  cat("\n\n")
}
```

## Dimension reduction

```{r tsne-umap, fig.wide = TRUE, fig.height = 8, fig.cap = "t-SNEs colored by cluster ID (a) and group ID (b)."}
df <- data.frame(reducedDim(sce, "TSNE"), colData(sce))
prettify <- list(
  labs(x = "t-SNE dim. 1", y = "t-SNE dim. 2"),
  theme(aspect.ratio = 1, legend.key.height = unit(2, "mm"), legend.position = "bottom"))

p1 <- ggplot(df, aes(x = tSNE_1, y = tSNE_2, col = cluster_id)) +
  scale_color_manual(values = muscat:::cluster_colors) +
  geom_point(size = 0.1, alpha = 0.2) + guides(color = guide_legend(
    ncol = 3, override.aes = list(size = 2, alpha = 1))) + 
  theme_bw() + prettify

p2 <- ggplot(df, aes(x = tSNE_1, y = tSNE_2, col = group_id)) +
  geom_point(size = 0.1, alpha = 0.2) + guides(color = guide_legend(
    ncol = 1, override.aes = list(size = 2, alpha = 1))) + 
  theme_bw() + prettify

ls <- list(get_legend(p1), get_legend(p2))
ps <- lapply(list(p1, p2), "+", theme(legend.position = "none"))
plot_grid(
  plot_grid(plotlist = ps, nrow = 1, labels = "auto"), 
  plot_grid(plotlist = ls, nrow = 1, rel_widths = c(3, 1)),
  nrow = 2, rel_heights = c(1, 0.2))
```

# Save SCE & unfiltered results
```{r}
saveRDS(sce, file.path("output", "MAGL-SCE.rds"))
saveRDS(res, file.path("output", "MAGL-DS_res.rds"))
```